Ionospheric Disturbances in Mexican Territory Produced by Objects Entering the Atmosphere from Space \tnoteref{tnote1}

### 2022-02-12 paper/appendix.tex
update of wavelet program
Now we can measure wavelet coherence spectrum. Need a .gitignore file
Up to date paper, journal readme and figures
Up to date:
- journal.org with work desription
- figures of paper
- paper draft for elsevier, MDPI and arXiv formats (I don't know yet in which jornal I will publish)
- Update README
- Updated python codes

Where $r$ is measured from the center of the earth, $ -180^\circ < phi <= 180^\circ$ represents the longitude; is positive at east of Greenwich meridian, and negative eastwards. $\-90^\circ <= \theta <= 90^\circ$ represents the latitude and is positive at the north of equator and negative southwards.

@@

Since $r_{GLM}$ and $r_{obj}$ are measured from the center of the arth ge find that:

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{../energy_fit}
  \caption{Left: Linear fit between energies calculated from GLM data and energies reported by USG sensors. Right: Residuals of the fit, used as error in the recallibration factor. The three events used for this linear fit are GLM-00 (the cuban meteoroid), GLM-23 and GLM-Ven (Venezolan meteoroid).}
  \label{fig:lin_fit}
\end{figure}

>>>>

Where $r$ is measured from the center of the earth, $ -180^\circ < \phi <= 180^\circ$ represents the longitude; is positive at east of Greenwich meridian, and negative eastwards. $\-90^\circ <= \theta <= 90^\circ$ represents the latitude and is positive at the north of equator and negative southwards.

@@

Since $r_{GLM}$ and $r_{obj}$ are measured from the center of the earth we find that:

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{../figures/energy_fit}
  \caption{Left: Linear fit between energies calculated from GLM data and energies reported by USG sensors. Right: Residuals of the fit, used as error in the recallibration factor. The three events used for this linear fit are GLM-00 (the cuban meteoroid), GLM-23 and GLM-Ven (Venezolan meteoroid).}
  \label{fig:lin_fit}
\end{figure}

### 2022-02-12 paper/arXiv/paper.tex
Up to date paper, journal readme and figures
Up to date:
- journal.org with work desription
- figures of paper
- paper draft for elsevier, MDPI and arXiv formats (I don't know yet in which jornal I will publish)
- Update README
- Updated python codes

\title{Art\'iculo}
\author{Jorge Tarango-Yong}

@@

Cuerpo del texto aqu\'i
\end{document}

>>>>

%% Loading LaTeX packages

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
\usepackage{latexsym}

\usepackage{url}
\usepackage{xcolor}
\usepackage{multirow}
\usepackage{hyperref}
\usepackage{rotating}
\usepackage{aas_macros}
\usepackage{pdflscape}
\usepackage{longtable}
\usepackage{amssymb,amsmath,amsthm}
\usepackage{siunitx}
\usepackage[utf8]{inputenc}
%\usepackage[spanish]{babel}
\usepackage{etoolbox}
\usepackage{fixltx2e}
\usepackage{natbib}
\usepackage[top=2.5cm, left=2.5cm, right=2.5cm, bottom=2.5cm]{geometry}

\title{Ionospheric Disturbances in Mexican Territory Produced by Objects Entering the Atmosphere from Space}
%\author{Jorge Tarango-Yong}

@@

%% Enter abstract here

\begin{abstract}
%%%
Earth's ionosphere may be perturbed by different phenomena, such as %earthquakes, tsunamis or even events caused by humans, like nuclear explosions and rockets launched to space. 
geomagnetic storms, solar flares with a common origin in our star, the Sun. In this work we will study the effects in the ionosphere caused by objects coming from outside the atmosphere (asteroids) inside Mexican territory. The Geostationary Lightning Mapper (GLM) started to record the passage of fireballs since 2017 and has recorded over 3000 events in the American continent. By the other hand, the Center for Near Earth Object Studies has data of 886 meteors detected around the globe by the United States Government sensors (USG). From these databases, we chose a sample of events and collected data and estimated meteor properties such as mass and energy released (if not available). Also we collected GPS data from UNAVCO network for the days each event was reported looking for vestiges of the meteors passage. With this work, we have observed that the vTEC signal
 of the ionosphere observed and calculated with GPS data from stations on Earth, is affected by the passage of these objects producing ionospheric perturbations with a permanence of a twice hours in average.  
%%%%
\end{abstract}

% Begin entering content here

\section{Introduction}
\label{sec1}
\input{../introduction}

\section{Observational data}
\label{sec:methodology}
\input{../methodology}

\section{Bolides physical parameters}
\label{sec:bolides}
\input{../meteor-properties}
\section{Ionospheric background and wavelet power spectrum}
\label{sec:vTEC-maps}

\input{../vTEC-maps}

\section{Discussion and conclusions}
\label{sec:discussion}
\input{../discussion}

\section{Acknowledgments}
The author Tarango-Yong thanks to DGAPA-UNAM by the posdoctoral fellowship, as well as the Laboratorio de Ciencias Geoespacilaes (LACIGE) from the Escuela Nacional de Estudios Superiores -- Morelia for the facilities provided to carry out the calculations of this paper. In addition, Guti\'errez-Zalapa wishes to thanks the
Consejo Nacional de Ciencias y Tecnolog\'ia (CONACyT) for the support in carrying out this research, he also wishes to thanks the Posgrado en Ciencias de la Tierra from the Universidad Nacional Aut\'onoma de M\'exico. Finally, M. Rodr\'iguez-Mart\'inez also is grateful for the DGAPA PAPIIT/PAPIME projects (IN118119/PE103419) to the computing support.
%\label{sec:acknowledgments}
%\input{Acknowledgments}
%% Bibliography
%% Author year style
\input{../appendix}  
\bibliographystyle{mnras}
%\biboptions{authoryear}
\bibliography{../bibliography}

\end{document}

### 2022-02-12 paper/introduction.tex
Up to date paper, journal readme and figures
Up to date:
- journal.org with work desription
- figures of paper
- paper draft for elsevier, MDPI and arXiv formats (I don't know yet in which jornal I will publish)
- Update README
- Updated python codes

The Earth's magnetic field represents a final obstacle to the Solar Wind (SW) flux. When descelerated and defected by a non colisional shock wave in the flux direction, generates a cavity known as magnetosphere \citep{Blanco-Cano:2004}. Since the Earth is embedded in this SW flux, is known that under adequated physical conditions (e.g magnetic reconnection) may exist some coupling between the magnetosphere and the Earth's ionosphere \citep{Zolesi:2014, Cnossen:2014}.

The Sun plays an important role in the physical proccesses that occur in the terrestrial magnetosphere-ionosphere system. When the SW interacts with the Earth's magnetosphere, particles may permeate the internal region via magnetic reconnection and penetrate to polar zones and generate boreal or austral auroras thus altering the system \citep{Vazquez:2016, Oka:2011}. By the other hand, the Extreme Ultraviolet Radiation (EUV) and X-rays coming from the Sun may interact with the neutral atmoshere via photoionization \citep{Vlasov:2010}. However, in both cases the final result is that the ionosphere's free electrons population is altered.

@@

In this work we will show a similar analysis for a sample of meteoroids detected in mexican territory by different methods. The first subsample consists in objects detected by the Geostationary Lightning Mapper (GLM) whose sizes are estimated between a few decimeters to meters in diameter \citep{GOODMAN:2013, Jenniskens:2018, Rumpf:2019}. The second subsample will consist in objects detected by ocular witnesses from the American Meteor Society and as comparisson we will include the morelian meteoroid reported in \citet{Sergeeva:2020} and the Chelyabinsk event \citet{Yang:2014}. The paper is arranged in the following way: \S \ref{sec:methodology} describes the samples of meteoroids as well of the properties that can be obtained from direct observations. Also describes the GPS data corresponding to the dates and locations where each object was located. \S \ref{sec:bolides} shows physical parameters of meteoroids obtained from the observed heights and energies. Finally, section \S \ref{sec:vTEC-maps} shows the vTEC maps and scintillation indices obtained from GPS observations.

>>>>

The Earth's magnetic field represents a final obstacle to the Solar Wind (SW) flux. When decelerated and defected by a non colisional shock wave in the flux direction, generates a cavity known as magnetosphere \citep{Blanco-Cano:2004}. Since the Earth is embedded in this SW flux, is known that under adequate physical conditions (e.g magnetic reconnection) may exist some coupling between the magnetosphere and the Earth's ionosphere \citep{Zolesi:2014, Cnossen:2014}.

The Sun plays an important role in the physical processes that occur in the terrestrial magnetosphere-ionosphere system. When the SW interacts with the Earth's magnetosphere, particles may permeate the internal region via magnetic reconnection and penetrate to polar zones and generate boreal or austral auroras thus altering the system \citep{Vazquez:2016, Oka:2011}. By the other hand, the Extreme Ultraviolet Radiation (EUV) and X-rays coming from the Sun may interact with the neutral atmoshere via photoionization \citep{Vlasov:2010}. However, in both cases the final result is that the ionosphere's free electrons population is altered.

@@

In this work we will show a similar analysis for a sample of meteoroids detected in Mexican territory by different methods. The first subsample consists in objects detected by the Geostationary Lightning Mapper (GLM) whose sizes are estimated between a few decimeters to meters in diameter \citep{GOODMAN:2013, Jenniskens:2018, Rumpf:2019}. The second subsample will consist in objects detected by ocular witnesses from the American Meteor Society and as comparisson we will include the morelian meteoroid reported in \citet{Sergeeva:2020} and the Chelyabinsk event \citet{Yang:2014}. The paper is arranged in the following way: \S \ref{sec:methodology} describes the samples of meteoroids as well of the properties that can be obtained from direct observations. Also describes the GPS data corresponding to the dates and locations where each object was located. \S \ref{sec:bolides} shows physical parameters of meteoroids obtained from the observed heights and energies. Finally, section \S \ref{sec:vTEC-maps} shows the vTEC maps and scintillation indices obtained from GPS observations.

### 2022-02-12 paper/meteor-properties.tex
Up to date paper, journal readme and figures
Up to date:
- journal.org with work desription
- figures of paper
- paper draft for elsevier, MDPI and arXiv formats (I don't know yet in which jornal I will publish)
- Update README
- Updated python codes

\subsection{Meteoroid masses}

Raul should write this section

>>>>

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{../figures/Energy_vs_Mass-fireballs_en}
    \caption{Bolides mass derived from Fragment Cloud Model. We used three assumptions about the bolide velocity: Minimum velocity $(\sim \SI{15}{km.s^{-1}})$, Medium velocity $(\sim \SI{16}{km.s^{-1}})$ and maximum velocity $(\sim \SI{18}{km.s^{-1}})$. This values are based on the available data of USG sample (see table \ref{tab:table-meteors-2})}
    \label{fig:energy-mass}
\end{figure}

Once the estimated the bolides energy, we are capable to estimate other bolides parameters, such as the mass. For that we used a Fragment Cloud Model \citep{Wheeler:2017}. To obtain the bolide mass, we must have previous knowledge of its velocity. Based on the few events of the USG sample where we could extract the velocity (see table \ref{tab:table-meteors-2}), we tried three models of bolides mass as a function of its releasing energy: the low velocity model, which assumes a velocity of $\sim \SI{15}{km.s{^-1}}$, medium velocty $v \sim \SI{16}{km.s{^-1}}$ and high velocity $v \sim \SI{18}{km.s{^-1}}$. The resultant masses can be seen in figure \ref{fig:energy-mass}. For the rest of the sample we need to assume that the bolides velocity is within this range or close enough in order to get a reasonable estimation of its mass.

### 2022-02-12 paper/methodology.tex
Up to date paper, journal readme and figures
Up to date:
- journal.org with work desription
- figures of paper
- paper draft for elsevier, MDPI and arXiv formats (I don't know yet in which jornal I will publish)
- Update README
- Updated python codes

By the other hand, we got another sample of 10 bolides from US Goverment (USG) sensors from the Center for Near Earth Object Studies (CNEOS), publicly available at \url{https://cneos.jpl.nasa.gov/fireballs/}, where we may obtain directly data about bolides position, the date and time each bolide was detected, the energy released at fragmentation, the velocity and the height (the last two not available for all bolides). As seen in table \ref{tab:table-meteors-2}, the time span is quite larger, and the released energy is generally larger. Some elements appear in both samples, since are bright enough to be detected regardless the project involved. In USG sample, the total energy of each meteor is already available from the database, but the energy for GLM bolides needs to be estimated. To do so, we estimated the distance between GOES satellites and the bolide and then followed \citet{Jenniskens:2018}, but we find some discrepancies in our measurements repect what is already reported for the same bolides in the USG database. (see the appendix \ref{app:distance} and \ref{app:energy}) for more details. We assumed that such discrepancy is due that GLM detectors underestimate the radiated energy by the bolides since before fragmentation the bolides remain undetectable by GLM sensors, and thus a fraction of the radiated energy is not accounted. To solve this, we estimated a correction factor using the bolides that appears in both GLM and USG samples (see figure \ref{fig:lin_fit} in appendix \ref{app:energies}). The energy distributions of USG and GLM samples are compared directly in figure \ref{fig:boxplot}.

@@

  % Both parameters sTEC and vTEC are computed
 Using a software developed by Gopi K. Seemala, publicly available at \url{https://seemala.blogspot.com/}, we computed the slant TEC (TEC) and vertical TEC (vTEC) for a several number of GPS satellites, each one identified with a PseudoRandom Noise code (PRN). An example of such TEC calculations is shown in figure \ref{fig:TEC-curve-example}. The behavior of the TEC curve is due to many factors, including the earth's rotation, solar activity, etc. TID's and wave-like features are not as prominent and are difficult to see. So we focused in a time interval close to the moment the bolide was detected, and detrended our time series following \citet{Pradipta:2015} using a Golay-Savitzky filter of order 7. An example of the resulting detrending signal is shown in figure . The resulting time series are adequate for wavelet analysis.%For each station, we filtered all the PRN and data that lie in a window of 30 minutes around the time the event was reported. Then we estimated the signal trend using a Savitsky-Golay filter and substracting the original signal from the trend.

\begin{figure}
\includegraphics[width=\linewidth]{../figures/abmf173-2019-06-22_page0}
\label{fig:TEC-curve-example}
\caption{Example of TEC curve obtained for GPS data for the bolide USG-09 from ABMF station.}
\end{figure}

\begin{figure}
\includegraphics[width=\linewidth]{}
\end{figure}

@@

>>>>

By the other hand, we got another sample of 10 bolides from US Goverment (USG) sensors from the Center for Near Earth Object Studies (CNEOS), publicly available at \url{https://cneos.jpl.nasa.gov/fireballs/}, where we may obtain directly data about bolides position, the date and time each bolide was detected, the energy released at fragmentation, the velocity and the height (the last two not available for all bolides). As seen in table \ref{tab:table-meteors-2}, the time span is quite larger, and the released energy is generally larger. Some elements appear in both samples, since are bright enough to be detected regardless the project involved. In USG sample, the total energy of each meteor is already available from the database, but the energy for GLM bolides needs to be estimated. To do so, we estimated the distance between GOES satellites and the bolide and then followed \citet{Jenniskens:2018}, but we find some discrepancies in our measurements repect what is already reported for the same bolides in the USG database. (see the appendix \ref{app:distance} and \ref{app:energy}) for more details. We assumed that such discrepancy is due that GLM detectors underestimate the radiated energy by the bolides since before fragmentation the bolides remain undetectable by GLM sensors, and thus a fraction of the radiated energy is not accounted. To solve this, we estimated a correction factor using the bolides that appears in both GLM and USG samples (see figure \ref{fig:lin_fit} in appendix \ref{app:energy}). The energy distributions of USG and GLM samples are compared directly in figure \ref{fig:boxplot}.

@@

  % Both parameters sTEC and vTEC are computed
 Using a software developed by Gopi K. Seemala, publicly available at \url{https://seemala.blogspot.com/}, we computed the slant TEC (TEC) and vertical TEC (vTEC) for a several number of GPS satellites, each one identified with a PseudoRandom Noise code (PRN). An example of such TEC calculations is shown in figure \ref{fig:TEC-curve-example}. The behavior of the TEC curve is due to many factors, including the earth's rotation, solar activity, etc. TID's and wave-like features are not as prominent and are difficult to see. So we focused in a time interval close to the moment the bolide was detected, and detrended our time series following \citet{Pradipta:2015} using a Savitzky-Golay filter of order 7 (future change to this phrase: to do a proper detrending, we should detrend data for a TEC series of a random day when no geomagnetic storm or unusual events happened, callibrate the Savitsky-Golay filter for using a proper order and use it for our data). An example of the resulting detrending signal is shown in figure \ref{fig:detrending-example}. The resulting time series are adequate for wavelet analysis.%For each station, we filtered all the PRN and data that lie in a window of 30 minutes around the time the event was reported. Then we estimated the signal trend using a Savitsky-Golay filter and substracting the original signal from the trend.

\begin{figure}
\includegraphics[width=\linewidth]{../figures/cn05173-2019-06-22_page0}
\label{fig:TEC-curve-example}
\caption{Example of TEC curve obtained for GPS data for the bolide USG-09 from CN05 station.}
\end{figure}

\begin{figure}
\begin{tabular}{cc}
\includegraphics[width=0.45\linewidth]{../figures/cn05-6-190}&
\includegraphics[width=0.45\linewidth]{../figures/cn05-TEC_series_PRN_6}
\end{tabular}
\label{fig:detrending-example}
\caption{Example of detrending process with USG-09 event. Left: Undetrended TEC series for station CN05 with PRN 6 (blue) versus Golay-Savitsky fit of order 7 (red dashed curve). Right: detrended curve resulting from substracting the Golay-Savitsky fit from the original signal.}
\end{figure}

@@

Wavelet analysis is a powerful tool for analyzing variations of power within a time series \citep{Torrence:1998}. For this task we used a Morlet wavelet $\Psi_0(\eta) = \pi^{-1/4}\exp(i\omega_0\eta)\exp(-\eta^2/2)$, where $\omega_0$ is a non-dimensional frequency, usually set to be 6 to satisfy the admissibity condition \citep{Farge:1992} and $\eta$ is a non-dimensional "time parameter".

With this, the wavelet transform is given by:

\begin{align}
W_n(s) = \sum_{k=0}^{N-1}\hat{x}_k\hat{\Psi}^*\left(s\omega_k\right)\exp(i\omega_k n\delta t)
\end{align}

Where the $*$ indicates the complex conjugate, $k=0, 1, \ldots, N-1$ is the frequency index, s is the scale, $\delta t$ is the time spacing between data, $\omega_k$ is the angular frequency defined as:

\begin{align}
  \omega_k = \left\lbrace
  \begin{array}{ccc}
    \frac{2\pi k}{N\delta t} & : & k\leq \frac{N}{2} \\
    -\frac{2\pi k}{N\delta t} & : & k > \frac{N}{2} 
  \end{array}
  \right.
\end{align}
And $\hat{\Psi}^*\left(s\omega_k\right)$ is the normalized wavelet function:

\begin{align}
  \hat{\Psi}^*\left(s\omega_k\right) = \left( \frac{2\pi s}{\delta t}\right)^{1/2} \hat{\Psi}_0\left(s\omega_k\right)
\end{align}

Since $\Psi(\eta)$ is a complex function, is expected that $W_n(s)$ is also complex. The transform can be divided into real and imaginary part. From the amplitude $|W_n(s)|$, from which we can obtain the power spectrum $|W_n(s)|^2$.

If two stations show similar power spectrums, we can be confident enough to compute the wavelet coherence of both time series, defined as \citep{Yang:2011}:

\begin{align}
  R^2_{1,2}=\frac{|W_{12}|^2}{\left(|W_1|^2|W_2|^2\right)}
\end{align}

where the $W_{12}= W_1W^*_2$ is the cross wavelet spectrum, $W_i$ is the wavelet transform of the time series and $W^*_i$ is the complex conjugate of $W_i$. Regions with high coherence indicate the precence of a TID with frequency $1/T$, where $T$ is the period of the region with high frequency.

### 2022-02-12 paper/vTEC-maps.tex
Up to date paper, journal readme and figures
Up to date:
- journal.org with work desription
- figures of paper
- paper draft for elsevier, MDPI and arXiv formats (I don't know yet in which jornal I will publish)
- Update README
- Updated python codes

Ionospheric perturbations also can take place due to space weather and geomagnetic storms. So, in order to discard such events we investigated the space weather in the day each event occured. In figure (name) we present the geomagnetic \textit{Kp} index for some events. We discarded events whose Kp index is equal or grater than 4 in the day of the event or shortly before. Also we present in figure (name) the vTEC perturbation maps for the same events in a three day series, centered in the event date. The estimated meteor trajectory, obtained from the GLM data is presented in black, continous line, while the linear fit to the GOES-16 and GOES-17 data are presented with the red dashed line, and work as boudary errors.

>>>>

For each TEC series like figure \ref{fig:detrending-example} we computed $|W_n(s)|^2$ to find any wave-like features in any of the TEC series, where $s$ is the wavelet scale (somehow is the analogous to the wavelength). The presence of this features allow us to trace the TEC perturbations produced by the meteor. Examples are shown in figure \ref{fig:power-spectrums}. Due to the fact we are working with finite time series, edge effects become important at the beginning or the end of the time series, and when the wavelet scale is similar or greater than the time series length. The limit when these errors become important is called Cone of Influence (COI), the COI is shown as a white dashed line in figure \ref{fig:power-spectrums}. At a closer look of this spectra is clear that in some cases the wavelet spectra in the previous day is more intense than the event day. This happens because no TEC perturbations was really detected, but instead noise appears as wavelets. This is an indicator that the TEC perturbances either propagate to other location or it is too faint to be detectable.

\begin{figure*}
  \begin{tabular}{cc}
  \includegraphics[width=0.5\linewidth]{../figures/cn05-PRN2-contour}  & \includegraphics[width=0.5\linewidth]{../figures/cn27-PRN2-contour}\\
  \includegraphics[width=0.5\linewidth]{../figures/crlr-PRN2-contour}  & \includegraphics[width=0.5\linewidth]{../figures/crse-PRN2-contour}
  \end{tabular}
  \label{fig:power-spectrums}
  \caption{Examples of power spectrums for event USG-09, PRN 2. Top left: Station CN05. Top Right: Station CN27. Bottom left: station CRLR. Bottom right: station CRSE. In each figure the left panel corresponds to the event day and the right panel corresponds to the previous day of the event. The white dashed lines correspond to the Cone of Influence, where all the wavelet spectrum below is subject to error.}
\end{figure*}
